### 2018-05-18 manuscript.rmd
edits through end of 3.1, did not rebuild pdf

Siphonophores are a diverse group of hydrozoans (Cnidaria) that are found at most depths of the ocean - from the surface, like the familiar Portuguese man of war, to the deep sea. They play important roles in ocean ecosystems, and are among the most abundant gelatinous predators. A previous phylogenetic study based on two ribosomal RNA genes provided insight into the internal relationships between major siphonophore groups, however there was little support for many deep relationships within the clade Codonophora. Here, we present a new siphonophore phylogeny based on new transcriptome data from 29 siphonophore species analyzed in combination with 14 publicly available genomic and transcriptomic datasets. We use this new phylogeny to reconstruct several traits that are central to siphonophore biology, including sexual system (monoecy vs. dioecy), gain and loss of zooid types, life history traits, and habitat. The phylogenetic relationships in this study are largely consistent with the previous phylogeny, but we find strong support for new clades within Codonophora that were previously unresolved. These results have important implications for trait evolution within Siphonophora, including favoring the hypothesis that monoecy arose at least twice.

@@

New data were analysed in conjunction with 14 publicly available datasets [@zapata2015phylogenomic; @Dunn:2013kw; @lehnert2012developing; @philippe2009phylogenomics; @fidler2014unique; @chapman2010dynamic; @sanders2014differential; @putnam2007sea; @sanders2015interspecific], with a total number of 43 species. Sequence assembly, annotation, homology evaluation, gene tree construction, parsing of genes trees to isolate orthologous sequences, and supermatrix construction were conducted with Agalma (commit 6bd9988, running BioLite commit 784edc6) [@Dunn:2013kw; @Guang202416].  This workflow integrates a variety of existing tools [@grabherr2011full; @altschul1990basic; @doi:10.1093/nar/30.7.1575; @sukumaran2010dendropy; @langmead2012fast; @li2011rsem; @talavera2007improvement; @li2009sequence; @katoh2013mafft] and new methods. Maximum likelihood (ML) analyses of the supermatrix were conducted with RAxML v 8.2.0 [@stamatakis2006raxml] and implemented via Agalma. Bayesian Inference (BI) analyses of the supermatrix were conducted using Phylobayes v. 1.7a-mpi [@lartillot2009phylobayes]. Sequence alignments, sampled and consensus trees, and voucher information are available in the git repository. Tree figures were rendered with ggtree [@Yu:2016fo].

@@

ML analyses were conducted on the unpartitioned supermatrix using the WAG+$\Gamma$ model of amino acid substitution, and bootstrap values were estimated using `r length(bootstrap_trees)` replicates. BI was conducted using two different CAT models, CAT-Poisson and CAT-GTR [@doi:10.1093/molbev/msh112]. Two independent MCMC chains were run under the CAT-GTR model, and eight independent MCMC chains were run under the CAT-Poisson model. The BI analyses did not converge (CAT-Poisson: maxdiff=1, meandiff=0.0185032; CAT-GTR: maxdiff=1, meandiff=0.045415). Only the results from the CAT-Poisson model are presented here (Fig. S11). Two chains (chain1 and chain3) were stuck in local maxima after 1405 and 4695 iterations and were excluded from the analyses. Visual inspection of the traces indicated that a burn in of `r burn_in` trees was sufficient for all CAT-Poisson runs. This left `r length( phylobayes_trees )` trees in the posterior.

To test the robustness of the results to model selection, we also used IQTree to implement other ML analyses using different models of substitution (GTR, JTT)  [@doi:10.1093/molbev/msu300; @doi:10.1093/molbev/msx281].  We used ModelFinder to assess model fit [@kalyaanamoorthy2017modelfinder].We also evaluated effects of method (ML vs. BI), model, and program by running a fixed-state WAG+$\Gamma$ model in Phylobayes (Fig. S12).

@@

The phylogenetic relationships recovered in this study received strong support across analysis methods (Fig. \@ref(phylogeny)A), with a couple of localized exceptions (Fig. \@ref(phylogeny)B and \@ref(bayesian)). The ML results were robust to model selection (Fig. \@ref(fig:supplementary-trees)). BI results were not robust to model selection: BI CAT models produced strongly supported topologies that differed from the ML topology (Fig. \@ref(phylogeny)B and \@ref(bayesian)), while the BI WAG+$\Gamma$ topology was consistent with the ML topology (Fig. \@ref(fig:BI-WAG)). Most clades are consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,423-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. We provisionally refer here to Pyrostephidae as the clade including *Rudjakovia* sp., although sampling of *Pyrostephos vanhoeffeni* is needed in order to determine if *Rudjakovia* sp. falls within Pyrostephidae or is sister to it. Within the clade that is sister to Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. \@ref(phylogeny)A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

In ML analyses Euphysonectae consists of two reciprocally monophyletic groups that we here provisionally refer to as Clade A and Group B (Fig. \@ref(phylogeny)A). In BI CAT-Poisson analyses, Group B is graded (Fig. \@ref(phylogeny)B). The presence of an involucrum, a fold around the base of the cnidoband [@totton1965synopsis], is a potential synapomorphy for Clade A. Species of Clade A also have a descending mantle canal within the nectophores (Fig. \@ref(fig:desc-mantle), \@ref(fig:mantle-BI)), a structure that is also present in some calycophorans. Members of Clade A are also monoecious (Fig. \@ref(reconstruct)). There is not a clear synapomorphy for Group B. Within Group B there is high support for the placement of *Erenna richardi* in ML analyses, but it is placed as sister to Clade A in BI CAT-Poisson analyses (Fig. \@ref(phylogeny)B). More taxon sampling will be required to determine the relationship of species within this clade.

@@

```{r supplementary-trees, echo=FALSE, message=FALSE, warning=FALSE, verbose= FALSE, fig.width=6.5, fig.height=7, fig.cap= "Phylogenies obtained using different models of evolution in IQTree to test the robustness of the results to model selection. We used ModelFinder [@kalyaanamoorthy2017modelfinder] to assess model fit and to test the robustness of the results to model selection, we also used the WAG and GTR model. Unlabeled nodes have support >0.99. A. JTT + Empirically counted frequencies from alignment + FreeRate model with 7 categories (best model fit). Likelihood:-8113694.922; AIC:16227609.9565 .  B. GTR + Optimized base frequencies by maximum-likelihood + Free rate model with 6 categories. Likelihood: -8133157.335; AIC:16266530.7277. C. WAG + Optimized base frequencies by maximum-likelihood + Free rate model with 6 categories. Likelihood: -8156043.772; AIC score: 16312303.6120 "}

>>>>

Siphonophores are a diverse group of hydrozoans (Cnidaria) that are found at most depths of the ocean - from the surface, like the familiar Portuguese man of war, to the deep sea. They play important roles in ocean ecosystems, and are among the most abundant gelatinous predators. A previous phylogenetic study based on two ribosomal RNA genes provided insight into the internal relationships between major siphonophore groups. There was, however, little support for many deep relationships within the clade Codonophora. Here, we present a new siphonophore phylogeny based on new transcriptome data from 29 siphonophore species analyzed in combination with 14 publicly available genomic and transcriptomic datasets. We use this new phylogeny to reconstruct several traits that are central to siphonophore biology, including sexual system (monoecy vs. dioecy), gain and loss of zooid types, life history traits, and habitat. The phylogenetic relationships in this study are largely consistent with the previous phylogeny, but we find strong support for new clades within Codonophora that were previously unresolved. These results have important implications for trait evolution within Siphonophora, including favoring the hypothesis that monoecy arose at least twice.

@@

New data were analysed in conjunction with 14 publicly available datasets [@zapata2015phylogenomic; @Dunn:2013kw; @lehnert2012developing; @philippe2009phylogenomics; @fidler2014unique; @chapman2010dynamic; @sanders2014differential; @putnam2007sea; @sanders2015interspecific], with a total number of 43 species. Sequence assembly, annotation, homology evaluation, gene tree construction, parsing of genes trees to isolate orthologous sequences, and supermatrix construction were conducted with Agalma (commit 6bd9988, running BioLite commit 784edc6) [@Dunn:2013kw; @Guang202416].  This workflow integrates a variety of existing tools [@grabherr2011full; @altschul1990basic; @doi:10.1093/nar/30.7.1575; @sukumaran2010dendropy; @langmead2012fast; @li2011rsem; @talavera2007improvement; @li2009sequence; @katoh2013mafft] and new methods.

@@

We used ModelFinder [@kalyaanamoorthy2017modelfinder], as implemented in IQTree v1.6.3 [@doi:10.1093/molbev/msu300], to assess relative model fit. ModelFinder selected XX based on XX. To assess the robustness of our results, we conducted phylogenetic analyses using multiple software programs, methods (Maximum likelihood (ML) and Bayesian Inference (BI)), and models (including the model selected by ModelFinder and several other commonly used models). Maximum likelihood (ML) analyses were conducted with RAxML v8.2.0 [@stamatakis2006raxml] and IQTree v1.6.3 [@doi:10.1093/molbev/msu300; @doi:10.1093/molbev/msx281]. Bayesian Inference (BI) were conducted with Phylobayes v. 1.7a-mpi [@lartillot2009phylobayes]. Sequence alignments, sampled and consensus trees, and voucher information are available in the git repository. Tree figures were rendered with ggtree [@Yu:2016fo].

RAxML ML analyses were conducted on the unpartitioned supermatrix using the WAG+$\Gamma$ model of amino acid substitution (Figure XX). RAxML bootstrap values were estimated using `r length(bootstrap_trees)` replicates. IQTree ML analyses were run under JTT + Empirically counted frequencies from alignment + FreeRate model with 7 categories, the best model identified by ModelFinder, and the commonly used model GTR + Optimized base frequencies + Free rate model with 6 categories (Figure XX).

BI was conducted in phylobayes using two different models: fixed-state WAG+$\Gamma$ (Fig. S12) and CAT-Poisson (Figure S11). Eight chains were run under the CAT-Poisson model. The WAG+$\Gamma$ runs (XX provide the same detail as is available for CAT poisson). The CAT-Poisson runs did not converge (maxdiff=1, meandiff=0.0185032). Closer inspection revealed that chain1 and chain3 were stuck in local maxima with low likelihood relative to other chains after 1405 and 4695 iterations. These two chains were excluded from the analyses, and the results pressented here are based on the remaining 6 CAT-Poisson chains. Visual inspection of the traces indicated that a burn in of `r burn_in` trees was sufficient for all CAT-Poisson runs. This left `r length( phylobayes_trees )` trees in the CAT-Poisson posterior.

@@

The phylogenetic relationships recovered in this study received strong support across analysis methods (Fig. \@ref(phylogeny)A), with a couple of localized exceptions (Fig. \@ref(phylogeny)B and \@ref(bayesian)). All of the ML analyses were congruent with each other, regardless of model and software used (Fig. \@ref(fig:supplementary-trees)). These ML results were also congruent with the Phylobayes BI WAG+$\Gamma$ analyses (Fig. \@ref(fig:BI-WAG)). The Phylobayes BI CAT-Poisson result (Fig. \@ref(bayesian)), however, had a strongly supported topology that differed (Fig. \@ref(phylogeny)B) from the ML topology in a couple localized regions as described below. The fact that the Phylobayes BI WAG+$\Gamma$ is consistent with the WAG (and other) ML analyses suggests that the different topology recovered in the Phylobayes BI CAT-Poisson analyses is due to the different model rather than different software or methods. Here we take the conservative approach of considering relationships that differ between the Phylobayes BI CAT-Poisson analyses and other analyses to be unresolved.

Most clades are consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,423-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. We provisionally refer here to Pyrostephidae as the clade including *Rudjakovia* sp., although sampling of *Pyrostephos vanhoeffeni* is needed in order to determine if *Rudjakovia* sp. falls within Pyrostephidae or is sister to it. Within the clade that is sister to Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. \@ref(phylogeny)A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

In ML analyses and BI WAG analyses, Euphysonectae consists of two reciprocally monophyletic groups that we here provisionally refer to as Clade A and Group B (Fig. \@ref(phylogeny)A). In BI CAT-Poisson analyses, Group B is paraphyletic (Fig. \@ref(phylogeny)B). The presence of an involucrum, a fold around the base of the cnidoband [@totton1965synopsis], is a potential synapomorphy for Clade A. Species of Clade A also have a descending mantle canal within the nectophores (Fig. \@ref(fig:desc-mantle), \@ref(fig:mantle-BI)), a structure that is also present in some calycophorans. Members of Clade A are also monoecious (Fig. \@ref(reconstruct)). There is not a clear synapomorphy for Group B. Within Group B there is high support for the placement of *Erenna richardi* in ML analyses (XX and BI WAG??), but it is placed as sister to Clade A in BI CAT-Poisson analyses (Fig. \@ref(phylogeny)B). More taxon sampling will be required to determine the relationship of species within this group.

@@

```{r supplementary-trees, echo=FALSE, message=FALSE, warning=FALSE, verbose= FALSE, fig.width=6.5, fig.height=7, fig.cap= "Phylogenies obtained using different models of evolution in IQTree. We used ModelFinder [@kalyaanamoorthy2017modelfinder] to identify which examined model as the best relative model fit. We also present results for the commonly used WAG and GTR models. Unlabeled nodes have support >0.99. A. JTT + Empirically counted frequencies from alignment + FreeRate model with 7 categories (the model identified by ModelFinder as having the best fit). Log likelihood:-8113694.922; AIC:16227609.9565 .  B. GTR + Optimized base frequencies by maximum-likelihood + Free rate model with 6 categories. Log likelihood: -8133157.335; AIC:16266530.7277. C. WAG + Optimized base frequencies by maximum-likelihood + Free rate model with 6 categories. Log likelihood: -8156043.772; AIC score: 16312303.6120 "}
